---
title: "Group Project 1"
output: github_document
---
## Loading packages/read data
```{r message = FALSE, warning=FALSE}
library(tidyverse)
library(here)

arthritis <- read_csv(here::here("data", "arthritis_clean.csv"))
```

## Summarize flexion before
```{r flexion_before, message=FALSE}
arthritis %>% 
  group_by(leg, Group) %>%
  summarise(min = min(Flexion_Before), Q1 = quantile(Flexion_Before, .25),
            median = median(Flexion_Before), Q3 = quantile(Flexion_Before, .75),
            max = max(Flexion_Before), std = sd(Rotation_Before)) %>% 
  knitr::kable()

arthritis %>% 
  ggplot(aes(x = Flexion_Before, color = leg))+
  geom_boxplot()+
  facet_grid(cols = vars(Group))+
  labs(title = "Flexion Before",
       x = "Degrees",
       y = NULL)+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
It looks like participants in the treatment group had greater mobility already than those in the control group. Worth testing to see if it's significant.

## Summarize Flexion After
```{r flexion_after, message=FALSE}
arthritis %>% 
  group_by(leg, Group) %>%
  summarise(min = min(Flexion_After), Q1 = quantile(Flexion_After, .25),
            median = median(Flexion_After), Q3 = quantile(Flexion_After, .75),
            max = max(Flexion_After), std = sd(Flexion_After)) %>% 
  knitr::kable()

arthritis %>% 
  ggplot(aes(x = Flexion_After, color = leg))+
  geom_boxplot()+
  facet_grid(cols = vars(Group))+
  labs(title = "Flexion After",
       x = "Degrees",
       y = NULL)+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

## summarize diff_flexion
```{r diff_flexion, message=FALSE}
arthritis %>% 
  group_by(leg, Group) %>% 
  summarise(min = min(diff_Flexion), Q1 = quantile(diff_Flexion, .25),
            median = median(diff_Flexion), Q3 = quantile(diff_Flexion, .75),
            max = max(diff_Flexion), std = sd(diff_Flexion)) %>% 
  knitr::kable()

arthritis %>% 
  ggplot(aes(x = diff_Flexion, color = leg))+
  geom_boxplot()+
  facet_grid(cols = vars(Group))+
  labs(title = "Difference in Flexion",
       x = "Degrees",
       y = NULL)+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
The treatment appears to have increase flexion mobility by 4 degrees in the left leg and 2.5 degrees in the right leg.


## Summarize Rotation Before
```{r rotation_before, message=FALSE}
arthritis %>% 
  group_by(leg, Group) %>% 
  summarise(min = min(Rotation_Before), Q1 = quantile(Rotation_Before, .25),
            median = median(Rotation_Before), Q3 = quantile(Rotation_Before, .75),
            max = max(Rotation_Before), std = sd(Rotation_Before)) %>% 
  knitr::kable()

arthritis %>% 
  ggplot(aes(x = Rotation_Before, color = leg))+
  geom_boxplot()+
  facet_grid(cols = vars(Group))+
  labs(title = "Rotation Before",
       x = "Degrees",
       y = NULL)+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
The median rotation for both hips and for treatment groups is similar.

## Summarize Rotation After
```{r rotation_after, message=FALSE}
arthritis %>% 
  group_by(leg, Group) %>% 
  summarise(min = min(Rotation_After), Q1 = quantile(Rotation_After, .25),
            median = median(Rotation_After), Q3 = quantile(Rotation_After, .75),
            max = max(Rotation_After), std = sd(Rotation_After)) %>% 
  knitr::kable()

arthritis %>% 
  ggplot(aes(x = Rotation_After, color = leg))+
  geom_boxplot()+
  facet_grid(cols = vars(Group))+
  labs(title = "Rotation After",
       x = "Degrees",
       y = NULL)+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
The treatment group appears to have increase roation in both hips compared to the control group.

## Summarize diff_rotation
```{r diff_rotation, message=FALSE}
arthritis %>% 
  group_by(leg, Group) %>% 
  summarise(min = min(diff_Rotation), Q1 = quantile(diff_Rotation, .25),
            median = median(diff_Rotation), Q3 = quantile(diff_Rotation, .75),
            max = max(diff_Rotation), std = sd(diff_Rotation)) %>% 
  knitr::kable()

arthritis %>% 
  ggplot(aes(x = diff_Rotation, color = leg))+
  geom_boxplot()+
  facet_grid(cols = vars(Group))+
  labs(title = "Difference in Rotation",
       x = "Degrees",
       y = NULL)+
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
treatment appears to have improved rotation in both legs more than in the control group. The difference in the right leg is quite a bit larger.



## Things to test
* Is there a significant difference between the treatment and control group's initial flexion/rotation
* Is there a significant difference in difference after intervention between the treatment groups in flexion/rotation
* Does leg matter

## The Test
* Mann-Whitney Test

### Assumptions
* One dependent variable (continuous or Ordinal)
  * We have a continuous variable (flexion/rotation)
* One independent variable (categorical, independent)
  * Treatment and Control, Right and Left Leg
* Independence of observations
  *measures are for each leg. Comparing left and right leg is likely not independent. I'm not sure about comparing between treatment/control
  * possible solution would be to average the rotation/flexion between each patient's legs
* Distributions have kind of similar shapes/spreads
  *can test medians
  * if not same shape/spread can only test for distribution
  
## Is there a difference between treatment group's inital flexion/rotation

clean data
```{r}
flex_crl <- arthritis$Flexion_Before[arthritis$Group =='Control']
flex_trt <- arthritis$Flexion_Before[arthritis$Group =='Treatment']

rot_crl <- arthritis$Rotation_Before[arthritis$Group =='Control']
rot_trt <- arthritis$Rotation_Before[arthritis$Group =='Treatment']
```

test flexion
```{r}
wilcox.test(flex_crl, flex_trt)

median(flex_crl)
median(flex_trt)
```
The Distribution between the treatment and control groups for their initial flexion is different

test rotation
```{r}
wilcox.test(rot_crl, rot_trt)

median(rot_crl)
median(rot_trt)
```
We cannot say that the distributions between the treatment and control groups for the patients' inital rotation are different.


## Is there a difference in flexion/rotation after treatment between the groups?

clean data
```{r}
flex_diff_ctr <- arthritis$diff_Flexion[arthritis$Group == "Control"]
flex_diff_trt <- arthritis$diff_Flexion[arthritis$Group == "Treatment"]

rot_diff_ctr <- arthritis$diff_Rotation[arthritis$Group == "Control"]
rot_diff_trt <- arthritis$diff_Rotation[arthritis$Group == "Treatment"]
```

test flexion
```{r}
wilcox.test(flex_diff_ctr, flex_diff_trt)

median(flex_diff_ctr)
median(flex_diff_trt)
```
The distributions of the difference in flexion between the control and treatment group are significantly different

test rotation
```{r}
wilcox.test(rot_diff_ctr, rot_diff_trt)

median(rot_diff_ctr)
median(rot_diff_trt)
```
The distributions for the difference in rotation between the control and treatment group are significantly different.

## Does Leg Matter

I think we can do this by treating it as paired data. We can make a vector of the difference in all the left legs and do the same for the right legs. To do this we will have to find out what leg belongs to what patient. I assume leg 1 and 2 belong to patient one. 3 and 4 to patient two and so on.

cleaning
```{r}
diff_flex_crt_l <- arthritis$diff_Flexion[arthritis$Group == "Control" & arthritis$leg == "Left"]
diff_flex_crt_r <- arthritis$diff_Flexion[arthritis$Group == "Control" & arthritis$leg == "Right"]

diff_flex__trt_l <- arthritis$diff_Flexion[arthritis$Group == "Treatment" & arthritis$leg == "Left"]
diff_flex_trt_r <- arthritis$diff_Flexion[arthritis$Group == "Treatment" & arthritis$leg == "Right"]

diff_rot_crt_l <- arthritis$diff_Rotation[arthritis$Group == "Control" & arthritis$leg == "Left"]
diff_rot_crt_r <- arthritis$diff_Rotation[arthritis$Group == "Control" & arthritis$leg == "Right"]

diff_rot_trt_l <- arthritis$diff_Rotation[arthritis$Group == "Treatment" & arthritis$leg == "Left"]
diff_rot_trt_r <- arthritis$diff_Rotation[arthritis$Group == "Treatment" & arthritis$leg == "Right"]
```


test flexion control
```{r}
wilcox.test(diff_flex_crt_l, diff_flex_crt_r)

median(diff_flex_crt_l)
median(diff_flex_crt_r)
```
We cannot conclude that the distributions for the difference in flexion for the control group is different between legs.

test flexion treatment
```{r}
wilcox.test(diff_flex__trt_l, diff_flex_trt_r)

median(diff_flex__trt_l)
median(diff_flex_trt_r)
```
We cannot conclude that the distributions for the difference in flexion for the treatment group is different between legs.

test rotation control
```{r}
wilcox.test(diff_rot_crt_l, diff_rot_crt_r)

median(diff_rot_crt_l)
median(diff_rot_crt_r)
```
We cannot conclude that the distributions for the difference in rotation for the control group is different between legs

test rotation treatment
```{r}
wilcox.test(diff_rot_trt_l, diff_rot_trt_r)

median(diff_rot_trt_l)
median(diff_rot_trt_r)
```
We cannot conclude that the distributions for the difference in rotation for the treatment group is different between legs

all of this said, the sample size may be too small to draw any meaningful conclusions about an suspected different between leg.
